---
title: "Plant-bacteria associations are phylogenetically structured in the phyllosphere"
author: "Geneviève Lajoie"
output: html_notebook
---

## Libraries

Loading libraries.

```{r Libraries, include=FALSE}
library(reshape2)
library(vegan)
library(ggplot2)
library(ape)
library(phyr)
library(heatmap3)
library(ggtree)
library(adephylo)
library(phylosignal)
library(phylobase)
library(phylofactor)
library(phytools)
library(doParallel)
library(plyr)
library(phyloseq)
library(dendextend)
```

<br />

## Data import and formatting
Importing environment with formatted datatables.

General dataset
```{r}
load('~/Data_analysis.RData')

# Description of the objects therein:

# asvt : Bacterial phylogeny of the 232 most abundant ASVs
# bac.10K, bac.10K.a : Taxonomic and functional annotations based on the metagenomic sequencing
# comb.tree : Host phylogeny
# kegg.cat : Functional classification of KEGG functional genes
# metadata : Metadata for each of 24 bacterial communities and hosts
# ps.new : ASV table and taxonomic annotations based on 16S sequencing

```

Formatting bacterial phylogeny
```{r}
# Make ultrametric
asvt2<-force.ultrametric(asvt)
# Make dichotomous (for downstream analyses)
asvt2<-multi2di(asvt2)
```

## Descriptive statistics

```{r}
# A) Functional composition: Average across samples
bac.10K.m<-reshape2::melt(as.matrix(bac.10K.a))

# Merge with kegg.cat
bac.10K.cat<-merge(bac.10K.m, kegg.cat, by.x='Var2', by.y='Kegg', all.x=T)
colnames(bac.10K.cat)[1:3]<-c('Kegg','samp','Freq')

# Calculate abundance of each C cat for each sample
sam_sum<-aggregate(Freq~samp+C+B+A, bac.10K.cat, sum)

# Mean across samples
sam_mean<-aggregate(Freq~C+B+A, sam_sum, mean)

# Creating datafile for plotting
samA<-aggregate(Freq~A, sam_mean, FUN=function(x) sum(x)/sum(sam_mean$Freq))
samA$cat<-'A'

samB<-aggregate(Freq~B, sam_mean, FUN=function(x) sum(x)/sum(sam_mean$Freq))
samB$cat<-'B'

samC<-aggregate(Freq~C, sam_mean, FUN=function(x) sum(x)/sum(sam_mean$Freq))
samC$cat<-'C'
colnames(samA)[1]<-colnames(samB)[1]<-colnames(samC)[1]<-'Fun'

# Adding color cat
samA$A<-samA$Fun
samB$A<-bac.10K.cat$A[match(samB$Fun,bac.10K.cat$B)]
samC$A<-bac.10K.cat$A[match(samC$Fun,bac.10K.cat$C)]

meta.bind<-rbind(samA,samB,samC)

# Transform data for plotting
comp.bind <- ddply(meta.bind, .(cat),
                   transform, init = 1- cumsum(Freq) + (0.5 * Freq))
comp.bind <- ddply(comp.bind, .(cat, A),
                   transform, pos = rev(Freq))  #1- cumsum(Freq) + (0.5 * Freq)
comp.bind <- ddply(comp.bind, .(cat),
                   transform, finpos = 1- cumsum(pos) + (0.5 * pos))  #1- cumsum(Freq) + (0.5 * Freq)
comp.bind <- ddply(comp.bind, .(cat, A),
                   transform, fin = rev(finpos))

# Samples aggregated / within or across treatments 
# CatC are not aligning with the Bs. Potentially just remove catC: not very informative.
ggplot(data=comp.bind, aes(x=cat, y=Freq, fill=A))+
  geom_bar(color='black', stat='identity', group="Fun", position='stack') +                    
  geom_text(data=subset(comp.bind, Freq > 0.02), aes(x=cat, y=fin, label=Fun), size=3) +
  theme_bw()+
  theme(legend.position="none")+
  labs(x='Functional categories', y='Relative abundance')

# B) Taxonomic composition: Average across samples

## 16S data
# Append OTU table with taxo table
otu16<-as.data.frame(t(ps.new@otu_table@.Data))
# Sum of each ASV freq across samples
otu.sum<-as.data.frame(rowSums(otu16))

# Append (average across samples)
taxo16<-as.data.frame(ps.new@tax_table@.Data)
taxo16$phylum[which(is.na(taxo16$phylum)==T)]<-'Unidentified phylum'
otu.taxo16<-merge(otu.sum,taxo16, by='row.names')
rownames(otu.taxo16)<-otu.taxo16$Row.names
otu.taxo16<-otu.taxo16[,-1]
colnames(otu.taxo16)[1]<-'Freq'

# Aggregate by phylum
phylo_sum<-aggregate(Freq~phylum, otu.taxo16, sum)
# Add unidentified bacteria
phylo_sum$Freq<-as.numeric(phylo_sum$Freq)
# Relative abundance
phylo_sum$Freq<-phylo_sum$Freq/sum(phylo_sum$Freq)

# Plot
ggplot(phylo_sum, aes(x=1, y=Freq, fill=phylum))+
  geom_bar(color='black', stat='identity',position='stack')+
  theme_bw()+
  labs(x='Host species', y='Relative abundance')

# Append (by sample)
otu16.m<-reshape2::melt(as.matrix(otu16))
otu16.m<-merge(otu16.m,taxo16, by.x='Var1',by.y='row.names', all.x=T)

# Aggregate across samples
otu16.samp<-aggregate(value~Var2+phylum, otu16.m, sum)
otu16.cast<-acast(otu16.samp, Var2~phylum, value.var='value')
otu16.cast<-decostand(otu16.cast, 'total')
otu16.samp<-reshape2::melt(otu16.cast)
otu16.samp$Var1<-as.character(otu16.samp$Var1)
otu16.samp$Var1<-metadata$Species[match(otu16.samp$Var1,metadata$sample_name)]

# Plot
ggplot(otu16.samp, aes(x=Var1, y=value, fill=Var2))+
  geom_bar(color='black', stat='identity',position='stack')+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90))+
  labs(x='Host species', y='Relative abundance')

## Metagenomic data
bac.taxo<-bac.10K
bac.taxo$X.5[which(is.na(bac.taxo$X.5)==T)]<-'Unidentified phylum'
bac.taxo$X.5[which(bac.taxo$X.5=='')]<-'Unidentified phylum'

# By sample
metax<-aggregate(seq1~sample_name+X.5, bac.taxo, length)
# Relative abundance
metax.cast<-acast(metax, sample_name~X.5, value.var='seq1', fill=0)
metax.cast<-decostand(metax.cast, 'total')
metax<-reshape2::melt(metax.cast)
metax$Var1<-as.character(metax$Var1)
metax$Var1<-metadata$Species[match(metax$Var1,metadata$sample_name)]

# Plot
ggplot(metax, aes(x=Var1, y=value, fill=Var2))+
  geom_bar(color='black', stat='identity',position='stack')+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90))

# Across samples
metax.all<-aggregate(value~Var2, metax, mean)

# Plot
ggplot(metax.all, aes(x=1, y=value, fill=Var2))+
  geom_bar(color='black', stat='identity',position='stack')

## Comparing taxonomic composition among 16S and metagenomic samples
# Remove unidentified samples and recompute relative abundances
metax.c<-metax.cast[,-which(colnames(metax.cast)=='Unidentified phylum')]
metax.c<-decostand(metax.c, 'total')
otu16.c<-otu16.cast[,-which(colnames(otu16.cast)=='Unidentified phylum')]
otu16.c<-decostand(otu16.c, 'total')

# Meta
meta.dist<-vegdist(metax.c)
met.pcoa<-pcoa(meta.dist)
# First dimensions
met.ps<-met.pcoa$vectors[,c(1:5)]

# 16S
otu.dist<-vegdist(otu16.c)
otu.pcoa<-pcoa(otu.dist)
# First dimensions
otu.ps<-otu.pcoa$vectors[,c(1:5)]

# Procrustes test
proct<-procrustes(met.pcoa$vectors[,1:12], otu.pcoa$vectors[,1:12], symmetric=T)
proct

funtest<-protest(X=met.pcoa$vectors[,1:12], Y=otu.pcoa$vectors[,1:12], scores = "sites", permutations = 999)
m12<-round(funtest$ss, digits=2)
corr<-round(funtest$t0, digits=2)
funtest

```

## Phylogenetic associations between plants and bacteria

Bacterial community composition among plant samples : formatting datasets
```{r}
# Prepare dataset (freq, sp, site)
dat<-as.data.frame(ps.new@otu_table)

# Replace site names by host species names
rownames(dat)<-metadata$Species[match(rownames(dat),metadata$sample_name)]
dat<-reshape2::melt(as.matrix(dat))
colnames(dat)<-c('site','sp','freq')

# Remove ASVs with <100 occurrence
dat<-dat[-which(dat$freq==0),] # Need to remove 0s
datagg<-aggregate(freq~sp, dat, sum)
dat2<-dat[-which(dat$sp%in%datagg[which(datagg$freq%in%c(1:99)),'sp']),] # 232 ASVs left (asvt phylogeny comprises these ASVs)
dat2$sp<-droplevels(dat2$sp)

# Proportion of the total ASV records that these ASV represent
sum(ps.new@otu_table@.Data[,which(colnames(ps.new@otu_table@.Data)%in%asvt2$tip.label)])/sum(ps.new@otu_table) # 77%

# Correct for unequal sample sizes
gg<-as.data.frame(acast(dat2, site~sp, value.var = 'freq', fill=0))
sort(rowSums(gg))
nef<-as.data.frame(rrarefy(gg, 3500)) #
dat3<-reshape2::melt(as.matrix(nef))
colnames(dat3)<-colnames(dat2)

# Matrix
dat.mat<-acast(dat3, site~sp, value.var='freq')
dat.hel<-decostand(dat.mat, 'hellinger')
```

Phylogenetic interaction model
```{r}
### Compute cophenetic distances for each phylogeny

## Bacterial phylogeny
newmic<-asvt2
# Variance cov matrix
VphyMic<-vcv(newmic)
VphyMic <- VphyMic[order(newmic$tip.label),order(newmic$tip.label)]

# Standardize (from pglmm function)
VphyMic <- VphyMic/max(VphyMic)
VphyMic <- VphyMic/exp(determinant(VphyMic)$modulus[1]/nrow(VphyMic))
det(VphyMic)

## Plant phylogeny
# Keep only species for which we have metagenomic data
plt.phy<-comb.tree
plt.phy<-drop.tip(plt.phy, setdiff(plt.phy$tip.label,rownames(dat.mat)))
plt.phy<-force.ultrametric(plt.phy)

VphyPlnt<-vcv(plt.phy)
VphyPlnt <- VphyPlnt[order(plt.phy$tip.label),order(plt.phy$tip.label)]

# Standardize (from pglmm function)
# Make it positive definite for chol decomposition to work
VphyPlnt <- VphyPlnt/max(VphyPlnt)
VphyPlnt <- VphyPlnt/exp(determinant(VphyPlnt)$modulus[1]/nrow(VphyPlnt))
det(VphyPlnt)

# General model of the form:

zf6.100 <- pglmm(log(freq+1) ~ 1 + (1|sp__) + (1|site__) + (1|sp__@site) + 
                  (1|sp@site__) + (1|sp__@site__), 
                data = dat3, family = "gaussian", # 
                cov_ranef = list(sp = VphyMic, site = VphyPlnt), 
                REML = TRUE, verbose = TRUE, s2.init = .1, maxit=100)

```

Likelihood ratio tests
```{r}
# Basic model
# Use summary function for variance components

za.100 <- glm(log(freq+1) ~ 1 , 
                data = dat3, family = "gaussian", # should be gaussian, not binomial
                maxit=100)

zal<-logLik(za.100)

# Random effect of species
zb.100 <- pglmm(log(freq+1) ~ 1 + (1|sp), 
                data = dat3, family = "gaussian", REML = TRUE, verbose = TRUE,# should be gaussian, not binomial
                maxit=100)

zbl<-zb.100$logLik

pvalue <- pchisq(2*(zbl - zal), df = 1, lower.tail = F)
pvalue

# Random effect of site
zc.100 <- pglmm(log(freq+1) ~ 1 + (1|sp) + (1|site), 
                 data = dat3, family = "gaussian", REML = TRUE, verbose = TRUE, # should be gaussian, not binomial
                 maxit=100)

zcl<-zc.100$logLik

pvalue <- pchisq(2*(zcl - zbl), df = 1, lower.tail = F)
pvalue

# Random effect of species phylogeny
zd.100 <- pglmm(log(freq+1) ~ 1 + (1|sp__) + (1|site), 
                data = dat3, family = "gaussian", cov_ranef = list(sp = VphyMic), REML = TRUE, verbose = TRUE, # should be gaussian, not binomial
                maxit=100)

zdl<-zd.100$logLik

pvalue <- pchisq(2*(zdl - zcl), df = 1, lower.tail = F)
pvalue

# Random effect of species phylogeny* site

ze.100 <- pglmm(log(freq+1) ~ 1 + (1|sp__) + (1|site) + (1|sp__@site), 
                data = dat3, family = "gaussian", cov_ranef = list(sp = VphyMic), REML = TRUE, verbose = TRUE, 
                s2.init = c(zd.100$ss, 0.01)^2,# should be gaussian, not binomial
                maxit=100)

zel<-ze.100$logLik

pvalue <- pchisq(2*(zel - zdl), df = 1, lower.tail = F)
pvalue

# Random effect of site phylogeny
zf.100 <- pglmm(log(freq+1) ~ 1 + (1|sp) + (1|site__), 
                data = dat3, family = "gaussian", cov_ranef = list(site = VphyPlnt), REML = TRUE, verbose = TRUE, # should be gaussian, not binomial
                maxit=100)

zfl<-zf.100$logLik

pvalue <- pchisq(2*(zfl - zcl), df = 1, lower.tail = F)
pvalue

# Random effect of site phylogeny* species
zg.100 <- pglmm(log(freq+1) ~ 1 + (1|sp) + (1|site__) + (1|site__@sp), 
                data = dat3, family = "gaussian", cov_ranef = list(site = VphyPlnt), REML = TRUE, verbose = TRUE, # should be gaussian, not binomial
                s2.init = c(zf.100$ss, 0.01)^2,
                maxit=100)

zgl<-zg.100$logLik

pvalue <- pchisq(2*(zgl - zfl), df = 1, lower.tail = F)
pvalue

# Random effect of site phylogeny & species phylogeny

zr6.100 <- pglmm(log(freq+1) ~ 1 + (1|sp__) + (1|site__) + (1|sp__@site) + # Can use any trait as sp*X instead of the intercept
                   (1|sp@site__), 
                 data = dat3, family = "gaussian", # should be gaussian, not binomial
                 cov_ranef = list(sp = VphyMic, site = VphyPlnt), 
                 REML = TRUE, verbose = TRUE, s2.init = .1, maxit=100)

zf6.100 <- pglmm(log(freq+1) ~ 1 + (1|sp__) + (1|site__) + (1|sp__@site) + # Can use any trait as sp*X instead of the intercept
                   (1|sp@site__) + (1|sp__@site__), 
                 data = dat3, family = "gaussian", # should be gaussian, not binomial
                 cov_ranef = list(sp = VphyMic, site = VphyPlnt), 
                 REML = TRUE, verbose = TRUE, s2.init = c(zr6.100$ss, 0.01)^2, maxit=100)

pvalue <- pchisq(2*(zf6.100$logLik - zr6.100$logLik), df = 1, lower.tail = F)
pvalue

```

Plot associations
```{r}
# Heatmap 
# https://stackoverflow.com/questions/15153202/how-to-create-a-heatmap-with-a-fixed-external-hierarchical-cluster
mata<-acast(dat3, site~sp, value.var = 'freq')
mata<-log(mata+1)

# Prepare association matrix
dat.plot<-dat3[-which(dat3$freq==0),]
dat.plot<-dat.plot[,c(2,1,3)]
match(as.character(unique(dat.plot$sp)),newmic$tip.label)
obj<-cophylo(asvt2, plt.phy, dat.plot[,c(1:2)])

# Ordering using basic co-phylogeny default (above)
pp<-as.dendrogram(force.ultrametric(obj[[1]][[2]], method='extend'))
mm<-as.dendrogram(obj[[1]][[1]])
mata<-mata[match(labels(pp),rownames(mata)),match(labels(mm),colnames(mata))]

# Use side colors to indicate taxonomy
# Subclass (4)
cols<-cbind(c('#99FF33','#33FF00','#006600','#003300'),c(1:4))
host.col.sc<-as.numeric(as.factor(metadata$Subclass[match(rownames(mata),metadata$Species)]))
host.col.sc<-cols[match(host.col.sc,cols[,2]),1]
# Order(7)
cols<-cbind(c('#99FF33','#669900','#006600','#CCFF00','#003300','#999933','#66CC66'),c(1:7))
host.col.ord<-as.numeric(as.factor(metadata$Order[match(rownames(mata),metadata$Species)]))
host.col.ord<-cols[match(host.col.ord,cols[,2]),1]

# Microbial taxonomy
cols2<-cbind(c("#CCCCCC","#000033","#000099","#0033FF","#0099FF","#FF99FF","#CC66FF","#99CCFF","#6666CC"),c(0:8))
cols3<-cbind(c("#CCCCCC","#73B761","#4A588A","#ECC846","#CD4C46","#71AFE2","#8D6FD1","#EE9E64","#95DABB"),c(0:8))
phyl<-as.data.frame(ps.new@tax_table@.Data)
# Identify NAs if possible
mtt<-phyl[match(colnames(mata),rownames(phyl)),]
mtt[which(is.na(mtt$class)==T),] # 
phyl$class[which(rownames(phyl)=='ASV10863')]<-phyl$class[which(rownames(phyl)=='ASV10154')]

# Match colors
mic.col.c<-as.numeric(as.factor(phyl$class[match(colnames(mata),rownames(phyl))]))
mic.col.c[which(is.na(mic.col.c)==T)]<-0
mic.col.c<-cols2[match(mic.col.c,as.numeric(cols2[,2])),1]

# Plot
heatmap3(mata, Rowv=pp, Colv=mm, col = colorRampPalette(c("ivory", "orange2", "firebrick3"))(10), scale='none', RowSideColors=host.col.ord, ColSideColors = mic.col.c, cexRow = 0.9)

```

## Phylogenetic factorization

```{r}
### Phylofactorization to identify bacterial clades that are associated with host taxonomy
# Check tutorials: https://github.com/reptalex/phylofactor
# https://dfzljdn9uc3pi.cloudfront.net/2017/2969/1/PhyloFactor_tutorial.html
# Especially for visualization

# OTU Phylogenetic tree
OTUTable<-dat.mat[,match(newmic$tip.label, colnames(dat.mat))]

# Plant trait
plt.taxo<-metadata[which(metadata$Species%in%rownames(OTUTable)),c('Order','Family','Genus','Species','Subclass')]
plt.taxo<-plt.taxo[match(rownames(OTUTable),plt.taxo$Species),]

OTUTable<-t(OTUTable)

# OTU Taxonomy
taxonomy<-as.data.frame(ps.new@tax_table@.Data)
taxonomy<-taxonomy[which(rownames(taxonomy)%in%rownames(OTUTable)),]
taxo<-data.frame(asv=rownames(taxonomy),taxo=paste(taxonomy$domain, taxonomy$phylum, taxonomy$class, taxonomy$order, taxonomy$family, taxonomy$genus, sep=';'))
taxo<-taxo[match(newmic$tip.label, taxo$asv),]

# Phylofactorization 
# Check at fraction explained variance

PF<- PhyloFactor(OTUTable,asvt2,plt.taxo$Order,nfactors=10) # 
summary(PF)
PF$models[[1]]

smry <- pf.summary(PF,taxo,factor=c(i))# i=1
# names(smry)
pf.tidy(smry)

# Visualization
# https://www.color-hex.com/color/268e66
gg <- pf.tree(PF, method='factors', factors=1:10, top.layer=F, colors=c("#26828e","#8e2682","#828e26","#264e8e","#268e66","#8e3226","#8e6626","#6dcd59","#cd596d","yellow"))
gg


### Phylofactorization to identify plant clades that are associated with bacterial taxonomy

# Abundance table
PLTTable<-as.data.frame(t(OTUTable))

# Phylogeny
plt.phy
PLTTable<-PLTTable[match(plt.phy$tip.label,rownames(PLTTable)),]

# Bacterial taxonomy
taxobac<-taxonomy
taxobac<-taxobac[match(colnames(PLTTable),rownames(taxobac)),]

# Issues with NA factor level, so replace with unknown for now
taxobac<-apply(taxobac, 2, as.character)
taxobac[is.na(taxobac)==T]<-'UNK'
taxobac<-as.data.table(taxobac)
rownames(taxobac)<-rownames(taxonomy)

# Plant taxonomy
taxoplt<-plt.taxo
taxoplt<-data.frame(plt=taxoplt$Species, taxo=paste(taxoplt$Order,taxoplt$Family,taxoplt$Genus, taxoplt$Species, sep=';'))
taxoplt<-taxoplt[match(plt.phy$tip.label,taxoplt$plt),]

table(taxobac$phylum)

# Phylofactorization
PF2<- PhyloFactor(PLTTable,plt.phy,taxobac$family, nfactors=10) summary(PF2)

smry <- pf.summary(PF2,taxoplt,factor=4)
#names(smry)
pf.tidy(smry)

# Visualization
# https://www.color-hex.com/color/268e66
gg2 <- pf.tree(PF2, method='factors', factors=1:10, top.layer=F, colors=c("#26828e","#8e2682","#828e26","#264e8e","#268e66","#8e3226","#8e6626","#6dcd59","#cd596d","yellow"))
gg2

```

## Principal coordinates analysis

Ordination of ASV composition - UNIFRAC
```{r}
# Find plant traits and bacterial traits most linked with bacterial phylogenetic structure
## Calculate UNIFRAC
physeq2 = phyloseq(otu_table(decostand(dat.mat,'hellinger'),taxa_are_rows=F), asvt2)

registerDoParallel(cores=10)
dat.dist.uni<-UniFrac(physeq2, weighted=F, normalized=TRUE, parallel=TRUE, fast=TRUE)

## Pcoa for envfit
pcoa.dat.uni <- cmdscale(dat.dist.uni, k=nrow(dat.hel)-1, eig=T)
# Pcoa to get at % variance
# pcoa.dat2 <- pcoa(dat.dist.uni, correction="lingoes")
# pcoa.dat2$values

```

Test of an effect of host taxonomy
```{r}
# Test for an effect of the structure
plot.str<-data.frame(metadata[,which(colnames(metadata)%in%c('Species','Genus','Family','Order','Subclass'))])
plot.str<-plot.str[match(labels(dat.dist.uni), plot.str$Species),]
    
# Model
perm.str<-adonis2(dat.dist.uni ~ Subclass + Order:Subclass , data = plot.str) # Cannot split it much more since we don't have levels below order for most orders

# Reporting model
perm.str
```

Plotting ordination
Use K1 and K2
```{r}
# Fit functional traits in there
rownames(bac.10K.a)<-metadata$Species[match(rownames(bac.10K.a),metadata$sample_name)]
bac.10K.o<-bac.10K.a[match(rownames(dat.mat),rownames(bac.10K.a)),]
efitC1<-envfit(pcoa.dat.uni, decostand(bac.10K.o, 'standardize'), choices=c(1)) 
efitC2<-envfit(pcoa.dat.uni, decostand(bac.10K.o,'standardize'), choices=c(2)) 

# # Which are most highly correlated
k1.names<-names(sort(efitC1$vectors$r[efitC1$vectors$pvals<0.05],decreasing=T))[1:10] # Can vary the number of variables here
k2.names<-names(sort(efitC2$vectors$r[efitC2$vectors$pvals<0.05],decreasing=T))[1:10]
# 
knames<-unique(c(na.omit(k1.names),na.omit(k2.names)))

# New fit
efitC<-envfit(pcoa.dat.uni, decostand(bac.10K.o[,which(colnames(bac.10K.o)%in%c(knames))],'standardize'), choices=c(1,2)) # ,k2.names

# Fit plant traits in here
# Transform plant traits
plt.trt.sd<-decostand(metadata[,c('SLA','P','Ca','Wood.dens')], 'standardize')
rownames(plt.trt.sd)<-metadata$Species
plt.trt.sd<-plt.trt.sd[match(rownames(dat.mat),rownames(plt.trt.sd)),]

efitP<-envfit(pcoa.dat.uni, plt.trt.sd, choices=c(1,2))

# Fit bacterial taxonomy
dat.mat.m<-melt(as.matrix(dat.mat))
dat.mat.m<-merge(dat.mat.m, taxonomy, by.x='Var2', by.y='row.names', all.x=T)
dat.tax<-aggregate(value~class+Var1, dat.mat.m, sum)
dat.tax<-acast(dat.tax, Var1~class, value.var="value")
dat.tax<-dat.tax[match(rownames(dat.mat),rownames(dat.tax)),]
dat.tax<-decostand(dat.tax, 'standardize')

efitT<-envfit(pcoa.dat.uni, dat.tax, choices=c(1,2))

# Plotting results
ordiplot(pcoa.dat.uni, type='text', display='site', choices=c(1,2), cex=0.8) # , type='text', display='site'
ordispider(pcoa.dat.uni, choices=c(1,2),groups, display = "sites", kind = "se", lty=c(1), conf = 0.95, alpha = 0.05, lwd = 1.5, label=TRUE)

ordiplot(pcoa.dat.uni, type='text', display='site', choices=c(1,2)) # , type='text', display='site'
plot(efitC, col="blue", cex = 0.9)
plot(efitP, col="red", cex = 0.9)

ordiplot(pcoa.dat.uni, type='text', display='site', choices=c(1,2)) # , type='text', display='site'
plot(efitT, col="blue", cex = 0.9)

```


## Phylogenetic Principal Component Analysis

```{r}
# PPCA to find the bacterial functions most associated with plant phylogenetic structure

# Build phylo4d object
trt.strip<-bac.10K.o
trt.strip<-trt.strip[match(plt.phy$tip.label, rownames(trt.strip)),]
colnames(trt.strip)<-gsub("\\s\\[.*", '', colnames(trt.strip))
colnames(trt.strip)<-gsub("\\s", "_", colnames(trt.strip))
colnames(trt.strip)<-gsub("-", "_", colnames(trt.strip))
colnames(trt.strip)<-gsub(",", "_", colnames(trt.strip))
plt.dat <- phylo4d(plt.phy, tip.data=trt.strip)

# PPCA
map.ppca <- ppca(plt.dat, center=TRUE, scale=TRUE, scannf=FALSE, nfposi=5, method="Abouheif") 
plot(map.ppca)

# Explore representations

## Dim 1
a <- map.ppca$c1[,1] # loadings on PC 1
names(a) <- row.names(map.ppca$c1)
highContrib <- a[a< quantile(a,0.01) | a>quantile(a,0.990)]

kk<-kegg.cat[which(kegg.cat$Kegg%in%names(highContrib)),]
kk<-merge(kk, as.data.frame(highContrib), by.x='Kegg', by.y='row.names', all.x=T)
kk$sign<-'Pos'
kk$sign[which(kk$highContrib<0)]<-'Neg'
kk<-kk[order(kk$sign,kk$B),]

kk1<-kk

no<-kk$Kegg[!duplicated(kk$Kegg)]

af<-trt.strip[match(rownames(map.ppca$li),rownames(trt.strip)), names(highContrib)]
af<-af[,match(no, colnames(af))]

# Add plant trait data
pt<-plt.trt.sd[match(rownames(af),rownames(plt.trt.sd)),]

datSel <- cbind.data.frame(af, map.ppca$ls[,1], pt)

temp <- phylo4d(plt.phy, datSel)
table.phylo4d(temp, ratio.tree=1/5, cex.label=0.7, cex.symbol=0.6, show.node.label=F, box=F)


## Dim 2
a <- map.ppca$c1[,2]
names(a) <- row.names(map.ppca$c1)
highContrib <- a[a< quantile(a,0.01) | a>quantile(a,0.990)]


kk<-kegg.cat[which(kegg.cat$Kegg%in%names(highContrib)),]
kk<-merge(kk, as.data.frame(highContrib), by.x='Kegg', by.y='row.names', all.x=T)
kk$sign<-'Pos'
kk$sign[which(kk$highContrib<0)]<-'Neg'
kk<-kk[order(kk$sign,kk$B),]

kk2<-kk

no<-kk$Kegg[!duplicated(kk$Kegg)]

af<-trt.strip[match(rownames(map.ppca$li),rownames(trt.strip)), names(highContrib)]
af<-af[,match(no, colnames(af))]

# Add plant trait data
pt<-plt.trt.sd[match(rownames(af),rownames(plt.trt.sd)),]

datSel <- cbind.data.frame(af, map.ppca$ls[,2], pt)

temp <- phylo4d(plt.phy, datSel)
table.phylo4d(temp, ratio.tree=1/5, cex.label=0.7, cex.symbol=0.6, show.node.label=F, box=F)

```

## Structure of host-bacteria trait associations

### Signal of plant traits in microbial phylogeny

```{r}
# A) Visualizing plant trait average per ASV

# Plant functional distance among bacteria
mica<-as.matrix(decostand(t(dat.mat), 'total'))

# Check sample order
plt.trt<-metadata[,c('SLA','P','Ca','Wood.dens')]
rownames(plt.trt)<-metadata$Species
plt.trt<-plt.trt[match(colnames(mica),rownames(plt.trt)),]

# Bind it with the means
resum<-mica %*% as.matrix(plt.trt)
resum<-resum[match(asvt2$tip.label, rownames(resum)),]

# Plot these values onto a bacterial phylogeny
X=as.matrix(resum)
rownames(X)<-resum$SampleID

circ <- ggtree::ggtree(asvt2, layout = "circular")
df<-data.frame(decostand(X, 'standardize'))

p1 <- gheatmap(circ, df, offset=0, width=.5, colnames_angle=95, colnames_offset_y = -3) +
    scale_y_continuous(expand = c(0.02,0)) +
    scale_fill_viridis_c(option="B", name="Plant traits")


# Boxplots

# Examine differences among classes and among families
dfm<-df
dfm$asv<-rownames(dfm)
dfm<-melt(dfm)
colnames(dfm)<-c('asv','trt','value')
dfm<-merge(dfm, taxonomy, by.x='asv', by.y='row.names', all.x=T)

# Class 
# Just those that have at least 5 observations / class
nm.plus<-names(table(taxonomy$class)[which(table(taxonomy$class)>=5)])

ggplot(dfm[which(dfm$class%in%nm.plus),], aes(x=class, y=value, fill=trt))+
  geom_boxplot()+
  theme_bw()+
  facet_wrap(~trt, ncol=1)+
  theme(axis.text.x = element_text(angle = 90, hjust=1))

# Add number of observations per class on the graph
table(taxonomy$class)

# Family
# Just those that have at least 5 observations / class
nm.plus<-names(table(taxonomy$family)[which(table(taxonomy$family)>=5)])
#nm.plus<-nm.plus[which(nm.plus%in%taxonomy$family[which(taxonomy$class=='Alphaproteobacteria')])]

# Reorder levels by Class
dfm2<-dfm[which(dfm$family%in%nm.plus),]
dfm2$family<-factor(dfm2$family, levels=c("Acidobacteriaceae (Subgroup 1)", "Microbacteriaceae", "Micromonosporaceae", "Acetobacteraceae", "Beijerinckiaceae", "Sphingomonadaceae", "Hymenobacteraceae", "Burkholderiaceae", "Enterobacteriaceae", "Pseudomonadaceae"))

ggplot(dfm2, aes(x=family, y=value, fill=trt))+
  geom_boxplot()+
  theme_bw()+
  facet_wrap(~trt, ncol=1)+
  theme(axis.text.x = element_text(angle = 90, hjust=1))

# Add number of observations per family on the graph
table(taxonomy$family)

# B) At which phylogenetic scale are they structured?

# Deal with negative edge lengths
asvt3<-asvt2
asvt3$edge.length[which(asvt3$edge.length<0)]<-0

mic.cwm<-mica %*% as.matrix(plt.trt)
mic.cwm<-mic.cwm[match(asvt2$tip.label,rownames(mic.cwm)),]
mic.cwm<-apply(mic.cwm,2,sqrt)

# Verify match between phylogeny and tip data
plot(match(asvt3$tip.label,rownames(resum)))
mic.cwm<-decostand(resum, 'standardize')

p42<-phylobase::phylo4d(x = asvt3, tip.data = mic.cwm)

par(mfrow=c(2,2))
for (i in names(tdata(p42))){
plt.trt.MoranCor4 <- phyloCorrelogram(p42, i, n.points = 100, ci.bs = 1000)
# columns 1 and 4 in the res matrix are respectively axis x and y
plot(plt.trt.MoranCor4, main=i)
}

```

### Signal of microbial traits in plant phylogeny

```{r}
# At which phylogenetic scale are they structured
bac.trt<-bac.10K.a
bac.trt.sub<-bac.trt[,which(colnames(bac.trt)%in%unique(kk1$Kegg))]# Do it both with kk1 and kk2 [contrib at 0.005]
bac.trt.sub<-sqrt(bac.trt.sub)
bac.trt.sub<<-bac.trt.sub[match(plt.phy$tip.label,rownames(bac.trt.sub)),]

p4<-phylobase::phylo4d(x = plt.phy, tip.data = bac.trt.sub) # [,which(colnames(bac.trt)%in%rownames(plt.pic[which(plt.pic$PIC.variance.P<0.05),]
p4@order<-'preorder'

# This object for extracting depth results
# ** Not deterministic, results will change a bit from time to time
res.list<-NULL

par(mfrow=c(4,4))
for (i in names(tdata(p4))){
bac.trt.MoranCor <- phyloCorrelogram(p4, i, n.points = 25, ci.bs = 1000)
# columns 1 and 4 in the res matrix are respectively axis x and y
plot(bac.trt.MoranCor, main=i)
fg<-as.data.frame(bac.trt.MoranCor$res)
fg$trt<-i
res.list<-rbind(res.list, fg)

}

```

### Trait-trait correlations in both partners phylogenies

```{r}
# Test the combinations of plant and microbial traits

# Bacterial traits
bac.trt<-bac.10K.a
bac.trt.sub<-bac.trt[,which(colnames(bac.trt)%in%c(unique(kk1$Kegg),unique(kk2$Kegg)))]#
bac.trt.sub<-sqrt(bac.trt.sub)
bac.trt.sub<<-bac.trt.sub[match(plt.phy$tip.label,rownames(bac.trt.sub)),]
# Plant traits
plt.trt.sd<-plt.trt.sd[match(plt.phy$tip.label,rownames(plt.trt.sd)),]

# Join
plt.mic<-cbind(plt.trt.sd,bac.trt.sub)

# Associations while controlling for phylogenetic relationships
pglsModel <- nlme::gls(K00174 ~ SLA + Ca + P + Wood.dens , plt.mic, correlation = corBrownian(phy=plt.phy), method = "ML") 
summary(pglsModel)

# Full model
pglsModel.0 <- nlme::gls(K00174 ~ SLA + Ca + P + Wood.dens , plt.mic, method = "ML") # Check this for all traits that are important for explaining microbial betadiv / seems to have issues with trait values = 0 # + Ca + P + Wood.dens
summary(pglsModel.0)

# Comparing models
anova(pglsModel.0,pglsModel)
anova(pglsModel.0,pglsModel)$AIC[2]-anova(pglsModel.0,pglsModel)$AIC[1]#  delta AIC
anova(pglsModel.0,pglsModel)$logLik[2]-anova(pglsModel.0,pglsModel)$logLik[1]# delta loglik

lrtest(pglsModel.0, pglsModel)

# For examination
# Detected relationships between bacterial and plant traits
plt.mic$Taxa<-metadata$Subclass[match(rownames(plt.mic),metadata$Species)]
ggplot(plt.mic, aes(x=SLA, y=K00995))+
  geom_point(aes(color=Taxa))+
  geom_smooth(method='lm', se=F)+
  geom_smooth(aes(group=Taxa, color=Taxa), method='lm', se=F)


```

